Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-03-12 analysis/paper/paper.qmd
editing intro

The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods. The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc. [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern]. Stemmed points are considered to be the first evidence of a suite of new technologies defining the Late Paleolithic period in this region [@seong2016eastern; @seong2008tanged]. This is related to the fact that stemmed points appear to originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did]. Despite the importance of stemmed points, only a few studies to date have examined their likely uses. Previous work has mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

The purpose of this study is to examine the possible uses of stemmed points in order to understand how stemmed points played a role in the technological transition. We use the tip cross-sectional area (TCSA) metric to infer potential variability in weapon-use strategies, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation]. We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these changed over time. Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses? As part of our discussion, we examine possible links between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period. In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to the socio-environmental circumstances they found themselves in.

>>>>

The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic, from the Early to the Late Paleolithic periods. The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc. [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern]. Stemmed points are considered to be the first evidence of a suite of new technologies defining the Late Paleolithic period in this region [@seong2016eastern; @seong2008tanged]. This is related to the fact that stemmed points appear to originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did]. Despite the importance of stemmed points, only a few studies to date have examined their likely uses. Previous work has mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

The purpose of this study is to examine the possible uses of stemmed points in order to understand what role stemmed points may have played in the technological transition from Early to the Late Paleolithic in Korea. We use the tip cross-sectional area (TCSA) metric to infer weapon-use strategies based on comparison with other archaeological and ethnographic assemblages, assuming that different ranges of TCSA values correspond to different weapon types [@lombard2021variation]. We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these changed over time in Korea. Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point uses? We examine possible links between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period. In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to their socio-environmental circumstances.
